---
title: "Dowloading and Manipulating the Li & Ito Datasets"
author: "Michelle Kang & Jessica Schaub"
date: "February 7, 2019"
output: github_document
---
# Introduction

Our project deals with data from two papers, [Li et al 2016](http://www.plantphysiol.org/content/172/2/1334.long) & [Ohashi-Ito et al 2010](http://www.plantcell.org/content/22/10/3461.long#sec-14). 
Each paper has deposited their data to GEO, so we use the `GEOquery` package to access the data. 
Then we perform some basic data manipulation to filter for the data of interest. 
At the end, we provide a final description of the data format and provide ready-to-use data frames for further analysis, as well as csv files.

## Downloading Necessary Packages

```{r}
library(GEOquery)
library(biomaRt)
library(tidyverse)
library(data.table)
library(reshape2)
library(limma)
library(data.table)
library(knitr)
```

# Data Download

## Li et al. 2016

The data are accessed from GEO using accession number [GSE77153](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE77153).
Expression data and metadata are saved as separate files, so they must be downloaded separately.
The expression data is saved as a data frame `li_expdata`.
The metadata is saved as a data frame `li_metadata`.

```{r}
# This code keeps us from re-downloading the data every time
# If it exists on our computer already, R will only load it
if (file.exists("GSE77153.Rdata")) {
    # if previously downloaded
    load("GSE77153.Rdata")
} else {
    # Get geo object that contains our data and phenotype information
    geo_obj <- getGEO("GSE77153", GSEMatrix = TRUE)
    geo_obj <- geo_obj[[1]]
    head(geo_obj)
    save(geo_obj, file = "GSE77153.Rdata")
}

# Get expression data
li_expdata <- exprs(geo_obj)

#Get metadata
li_metadata <- pData(geo_obj)[, c("organism_ch1", "title", colnames(pData(geo_obj))[grep("characteristics", 
    colnames(pData(geo_obj)))])]

#Changing column names to something more informative
colnames(li_metadata) = c("organism", "sample_name", "ecotype", "vector", "tissue", "treatment", "age")
li_metadata$ecotype = as.factor(gsub("ecotype/background: ", "", li_metadata$ecotype))
li_metadata$vector = as.factor(gsub("vector: ", "",li_metadata$vector))
li_metadata$tissue = as.factor(gsub("tissue: ", "", li_metadata$tissue))
li_metadata$treatment = as.factor(gsub("treatment: ", "", li_metadata$treatment))
li_metadata$age = gsub("time after induction: ", "", li_metadata$age)

```

The metadata and expression data files are now in a table format. We need to convert them to data frames and do a bit more tyding up. We end by gathering the data into tidy form as `tidy_li_expdata`.

```{r}
#Convert to data frame
li_expdata <- data.frame(li_expdata)

#Set gene names as a variable, not row names
#Expression data
setDT(li_expdata, keep.rownames = TRUE)
colnames(li_expdata)[colnames(li_expdata)=="rn"] <- "gene_id"
#Metadata
setDT(li_metadata, keep.rownames = TRUE)
colnames(li_metadata)[colnames(li_metadata)=='rn'] <- "sample"

#Convert to long form
tidy_li_expdata <- li_expdata %>% 
  gather(key="sample", value="expression", 2:ncol(li_expdata))
```

## Ohashi-Ito et al. 2010

As above, we access the data from GEO with accession number [GSE20586](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM517080).
Again, expression data and metadata are saved as two different files so they need to be downloaded individually.
The expression data is saved as `ito_expdata` and the metadata is saved as `ito_metadata`.

```{r}
if (file.exists("GSE20586.Rdata")) {
    # if previously downloaded
    load("GSE20586.Rdata")
} else {
    # Get geo object that contains our data and phenotype information
    geo_obj <- getGEO("GSE20586", GSEMatrix = TRUE)
    geo_obj <- geo_obj[[1]]
    head(geo_obj)
    save(geo_obj, file = "GSE20586.Rdata")
}

# Get expression data
ito_expdata <- exprs(geo_obj)


ito_metadata <- pData(geo_obj)[, c("organism_ch1", "title", colnames(pData(geo_obj))[grep("characteristics", 
    colnames(pData(geo_obj)))])]

colnames(ito_metadata) = c("organism", "sample_name", "treatment", "age")
ito_metadata$treatment = as.factor(gsub("cultured cell type: ", "", ito_metadata$treatment))
ito_metadata$age = gsub("time point: ", "", ito_metadata$age)

```

As above, we need to convert the tables to data frames and perform some other tidying tasks.
The data is gathered at the end to tidy form for future analysis as `tidy_ito_expdata`.

```{r}
#Convert to data frame
ito_expdata <- data.frame(ito_expdata)

#Set gene names as a variable, not row names
#Expression Data
setDT(ito_expdata, keep.rownames = TRUE)
colnames(ito_expdata)[colnames(ito_expdata)=="rn"] <- "gene_id"
#Metadata
setDT(ito_metadata, keep.rownames = TRUE)
colnames(ito_metadata)[colnames(ito_metadata)=="rn"] <- "sample"

#Convert to long form
tidy_ito_expdata <- ito_expdata %>% 
  gather(key="sample", value="expression", 2:ncol(ito_expdata))
```

#Combining Metadata and Data 

The individual files for metadata and expression data can be combined by sample ID.

## Li et al. 2016

```{r}
full_li_data <- left_join(tidy_li_expdata, li_metadata, by = 'sample')
```


## Ohashi et al. 2010

```{r}
full_ito_data <- left_join(tidy_ito_expdata, ito_metadata, by = 'sample')
```


# Filtering Data

To make the data frames easier to work with, we will remove some of the unused columns and samples that we are not interested in.

## Li et al. 2016

We are only interested in samples with age 0 hrs or 12 hrs.

```{r}
# Remove some unused columns and filter for ages of interest
li_data <- full_li_data %>% 
  select(-organism, -sample_name, -ecotype, -tissue, -vector) %>% 
  filter(age == '0h'|age == '12h')
```


##Ohashi-Ito et al. 2010

```{r}
# Remove some unused columns
ito_data <- full_ito_data %>% 
  select(-organism, -sample_name)
```

# Data Description

At this point, both data sets are clean and they have the same information in the same format:

```{r}
head(li_data, 10) %>% 
  kable()

head(ito_data, 10) %>% 
  kable()
```

These data sets are stored as data frames under `li_data` and `ito_data` for analysis.
We will also export them as csv files.

```{r}
write_csv(li_data, "li_data.csv")
write_csv(ito_data, "ito_data.csv")
```

The variables are as follows:

- *gene_id* : Probe label
- *sample* : Sample id that the RNA came from
- *expression* : normalized values for expression, but each paper normalized differently so we will deal with this during the analysis step
- *treatment* : information about which cell lines and conditions the samples were taken from
- *age* : the age of the sample in hours, where time 0 is usually the addition of DEX or estrogen
